Effect of quality-adjusted production function inputs

### 2015-05-27 Articles/Presentation_ESEE_2015/ESEE_2015.Rnw

<<Models>>=
flavors <- c("unadjusted", "adjusted")
nests <- list(kle=c(1,2,3), lek=c(2,3,1), ekl=c(3,1,2))
model.types <- c("CD", "CES")

@@

<<Fitted_and_Residuals>>=
fitted_and_resid_data <- leaf_apply(models, class=c("CDEmodel", "cesModel", "LINEXmodel"), f = function(mod, id){
  df <- select(mod$data, Year, iGDP, Country)
  # Trim fitted to be same length as df to eliminate problems with possible NAs at end of fitted
  df$fitted <- yhat(mod)[c(1:length(df$Year))]
  df$Residuals <- resid(mod)[c(1:length(df$Year))]
  # Parse the id to find out whether this is quality-adjusted data
  id_list <- strsplit(id, split = "[.]")[[1]]
  if (id_list[[2]] == "unadjusted"){
    df$flavor <- "Unadjusted"
  } else if (id_list[[2]] == "adjusted"){
    df$flavor <- "Quality-adjusted"
  } else {
    stop("Unknown adjustment.")
  }
  if (id_list[[3]] == "noE"){
    df$energy <- "Without energy"
  } else if (id_list[[3]] == "withE"){
    df$energy <- "With energy"
  } else {
    stop("Unknown energy.")
  }
  df$model <- id_list[[4]]
  df$nest <- if (length(id_list) > 4) id_list[[5]] else NA
  return(df)  
}) %>% do.call(rbind, .)

# Relevel factors before plotting
fitted_and_resid_data$Country <- relevelFactor(as.factor(fitted_and_resid_data$Country), c("PT", "UK"))
fitted_and_resid_data$flavor <- relevelFactor(as.factor(fitted_and_resid_data$flavor), c("Unadjusted", "Quality-adjusted"))
fitted_and_resid_data$energy <- relevelFactor(as.factor(fitted_and_resid_data$energy), c("Without energy", "With Energy"))
@

@@

<<Fitted and Residual Graphs>>=
fittedGraph <- function(data){
  ggplot() +
    geom_line(data=filter(data, energy=="Without energy"), mapping=aes(x=Year, y=iGDP)) +
    geom_line(data=data, mapping=aes(x=Year, y=fitted, linetype=energy)) +
    scale_linetype_manual(values=c(2,3), name="") +
    facet_grid(Country ~ flavor) +
    scale_x_continuous(breaks=c(1960, 1980, 2000)) +
    scale_y_continuous(breaks=c(1, 2, 4, 6, 8, 10, 12)) +
    ylab("Indexed GDP (1960=1)") +
    xy_theme()
}

@@

<<Dynamic Solow Residuals>>=
# Calculate finite difference approximations of time derivatives for output and all factors of production
lag <- 3       # Desired lag for calculations. Set to n for differences across n years.
pad <- "tail"  # Desired padding for NA values

dframes <- list()
for (Coun in countries){
  for (flav in flavors){
    forms <- list(noE = keep_summands(formulas[[Coun]][[flav]], -3), withE = formulas[[Coun]][[flav]])
    for (form in forms){
      denom <- summands(form, n=length(summands(form))) # time variable
      nums <- all.vars(form) [1:(length(all.vars(form))-1)] # y, k, l, and e (if present)
      df <- historicalData[[Coun]][c(denom, nums)] # columns are: iYear, y, k, l, and e (if present)
      # Calculate names for the columns of finite difference derivatives
      derivNames <- lapply(1:length(nums), function(i){paste0("d", nums[[i]], "/", "dt")}) %>% unlist
      derivCols <- lapply(1:length(nums), function(i){
        # Calculate finite difference derivative columns
        ediff(df[[i+1]], pad=pad, lag=lag) / ediff(df[[1]], pad=pad, lag=lag)
      }) %>% do.call(cbind, .) %>% as.data.frame
      names(derivCols) <- derivNames
      df <- cbind(derivCols, historicalData[[Coun]][ , c(nums, "Year")])
      e <- if (length(summands(form))==4) "withE" else "noE"
      for (m in model.types){
        origCoeffs <- filter(AllCoeffs, model==m & energy==e & Country==Coun & flavor==flav)
        df <- cbind(df, origCoeffs)
        if (m == "CD"){
          lambda <- lapply(1:length(derivNames), function(i){
            # derivative term divided by its numerator.
            # derivative term is in the ith column of df
            # numerator is in i + length(derivNames) th column of df
            out <- df[[i]] / df[[i+length(derivNames)]]
            if (i > 1){
              # Need to make it negative and multiply by the corresponding alpha term.
              # Typical setup is that columns 2 is capital, and we need alpha_1 for capital.
              # So, subtract 1 from i.
              out <- - out * df[[paste0("alpha_", i-1)]]
            }
            return(out)
          }) %>% as.data.frame %>% rowSums
          df$lambda_fd <- lambda
          avg.lambda_fd <- mean(df$lambda_fd, na.rm = TRUE)
          # Fraction of growth explained by Solow Residual
          df$solowFrac <- df$lambda_fd / (df[[1]] / df[[length(derivNames) + 1]])
          avg.solowFrac <- mean(df$solowFrac, na.rm = TRUE)
        } else if (m == "CES"){
          if (e == "noE"){
            # kl nesting
#             df$lambda_fd <- df[[1]] / 
          } else if (e == "withE"){

@@

        } else {
          stop(paste("Unknown model", model))
        }
        df$flavor <- if (flav == "adjusted") "Quality-adjusted" else "Unadjusted"
        df$flavor <- relevelFactor(as.factor(df$flavor), c("Unadjusted", "Quality-adjusted"))
        df$energy <- if (e == "noE") "Without energy" else "With energy"
        df$energy <- relevelFactor(as.factor(df$energy), c("Without energy", "With energy"))
        dframes[[length(dframes)+1]] <- df
      }
    }
  }      
}
solowResiduals <- do.call(plyr::rbind.fill, dframes)
@

@@

<<DynamicSR_CD_Graph, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Dynamic Solow residuals for the Cobb-Douglas model.">>=
data <- filter(solowResiduals, model=="CD")
ggplot(data=data) +
  geom_line(mapping=aes(x=Year, y=lambda_fd, linetype=energy)) +
  scale_linetype_manual(values=c(2,3), name="") +
  facet_grid(Country ~ flavor) +
  scale_x_continuous(breaks=c(1960, 1980, 2000)) +
  xlab("") + 
  ylab(expression(lambda[ij])) +
  xy_theme()
@

>>>>

<<Models>>=
flavors <- c("unadjusted", "adjusted")
nests <- list(kle=c(1,2,3), lek=c(2,3,1), ekl=c(3,1,2))
model.types <- c("CD")
# model.types <- c("CD", "CES")

@@

<<Fitted_and_Residuals>>=
fitted_and_resid_data <- leaf_apply(models, class=c("CDEmodel", "cesModel", "LINEXmodel"), f = function(mod, id){
  df <- select(mod$data, Year, iGDP, Country)
  # Trim fitted to be same length as df to eliminate problems with possible NAs at end of fitted
  df$fitted <- yhat(mod)[c(1:length(df$Year))]
  df$Residuals <- resid(mod)[c(1:length(df$Year))]
  # Parse the id to find out whether this is quality-adjusted data
  id_list <- strsplit(id, split = "[.]")[[1]]
  df$flavor <- if (id_list[[2]] == "unadjusted") "Unadjusted" else "Quality-adjusted"
  df$energy <- if (id_list[[3]] == "noE") "Without energy" else "With energy"
  df$model <- id_list[[4]]
  df$nest <- if (length(id_list) > 4) id_list[[5]] else NA
  return(df)  
}) %>% do.call(rbind, .)

# Get the unadjusted, without-energy residuals and store them in a data frame
r_unadj_noE <- filter(fitted_and_resid_data, flavor=="Unadjusted" & energy=="Without energy") %>% 
               select(Year, Country, Residuals, model)
colnames(r_unadj_noE)[colnames(r_unadj_noE)=="Residuals"] <- "r_unadj_noE"
# Merge to get the unadjusted, without-energy values next to each country, year, and model
fitted_and_resid_data <- merge(fitted_and_resid_data, r_unadj_noE)
# Create a column containing ∆r, the difference between a residual value
# and the residual value in the same year for the unadjusted, no-energy case
fitted_and_resid_data[["Dr"]] <- abs(fitted_and_resid_data[["Residuals"]]) - abs(fitted_and_resid_data[["r_unadj_noE"]])

# Relevel factors before plotting
fitted_and_resid_data$Country <- relevelFactor(as.factor(fitted_and_resid_data$Country), c("PT", "UK"))
fitted_and_resid_data$flavor <- relevelFactor(as.factor(fitted_and_resid_data$flavor), c("Unadjusted", "Quality-adjusted"))
fitted_and_resid_data$energy <- relevelFactor(as.factor(fitted_and_resid_data$energy), c("Without energy", "With energy"))
@

@@

<<Fitted and Residual Graph Functions>>=
fittedGraph <- function(data){
  ggplot() +
    geom_line(data=filter(data, energy=="Without energy"), mapping=aes(x=Year, y=iGDP)) +
    geom_line(data=data, mapping=aes(x=Year, y=fitted, linetype=energy)) +
    scale_linetype_manual(values=c(2,3), name="") +
    facet_grid(Country ~ flavor) +
    scale_x_continuous(breaks=c(1960, 1980, 2000)) +
    scale_y_continuous(breaks=c(1, 2, 4, 6, 8, 10, 12)) +
    ylab("Indexed GDP (1960=1)") +
    xy_theme()
}

@@

<<Dynamic Solow Residuals>>=
# Calculate finite difference approximations of time derivatives for output and all factors of production
lag <- 1       # Desired lag for calculations. Set to n for differences across n years.
pad <- "tail"  # Desired padding for NA values

dframes <- list()
for (Coun in countries){
  for (flav in flavors){
    forms <- list(noE = keep_summands(formulas[[Coun]][[flav]], -3), withE = formulas[[Coun]][[flav]])
    for (form in forms){
      denom <- summands(form, n=length(summands(form))) # time variable
      nums <- all.vars(form) [1:(length(all.vars(form))-1)] # y, k, l, and e (if present)
      df <- historicalData[[Coun]][c(denom, nums)] # columns are: iYear, y, k, l, and e (if present)
      # Calculate names for the columns of finite difference derivatives
      derivNames <- lapply(1:length(nums), function(i){paste0("d", nums[[i]], "/", "dt")}) %>% unlist
      derivCols <- lapply(1:length(nums), function(i){
        # Calculate finite difference derivative columns
        ediff(df[[i+1]], pad=pad, lag=lag) / ediff(df[[1]], pad=pad, lag=lag)
      }) %>% do.call(cbind, .) %>% as.data.frame
      names(derivCols) <- derivNames
      df <- cbind(derivCols, historicalData[[Coun]][ , c(nums, "Year")])
      e <- if (length(summands(form))==4) "withE" else "noE"
      df$flavor <- if (flav == "unadjusted") "Unadjusted" else "Quality-adjusted"
      df$energy <- if (e == "noE") "Without energy" else "With energy"
      # Add a column for 1/y * dy/dt
#       growth.col.name <- paste0("1/", nums[[1]], "*", derivNames[[1]])
      df[["dydtOvery"]] <- 1 / df[[nums[[1]]]] * df[[derivNames[[1]]]]
      for (m in model.types){
        origCoeffs <- filter(AllCoeffs, model==m & energy==e & Country==Coun & flavor==flav)
        df <- cbind(df, origCoeffs)
        if (m == "CD"){
          lambda_fd <- lapply(1:length(derivNames), function(i){
            # derivative term divided by its numerator.
            # derivative term is in the ith column of df
            # numerator is in i + length(derivNames) th column of df
            out <- df[[i]] / df[[i+length(derivNames)]]
            if (i > 1){
              # Need to make it negative and multiply by the corresponding alpha term.
              # Typical setup is that columns 2 is capital, and we need alpha_1 for capital.
              # So, subtract 1 from i.
              out <- - out * df[[paste0("alpha_", i-1)]]
            }
            return(out)
          }) %>% as.data.frame %>% rowSums
          df$lambda_fd <- lambda_fd
          avg.lambda_fd <- mean(df$lambda_fd, na.rm = TRUE)
          # Fraction of growth explained by Solow Residual
          df$solowFrac <- df$lambda_fd / (df[[1]] / df[[length(derivNames) + 1]])
          avg.solowFrac <- mean(df$solowFrac, na.rm = TRUE)
        } else if (m == "CES"){
          df <- data.frame()
          if (e == "noE"){
            # kl nesting
#             df$lambda_fd <- df[[1]] / 
          } else if (e == "withE"){

@@

        } else {
          stop(paste("Unknown model", model))
        }
        # Add to the list of data frames
        dframes[[length(dframes)+1]] <- df
      }
    }
  }      
}
solowResiduals <- do.call(plyr::rbind.fill, dframes)

# Grab the unadjusted, no-energy Solow residuals and create a column in solowResiduals with them. 
# Using the hint at 
# http://stackoverflow.com/questions/15059076/
# r-how-to-call-apply-like-function-on-each-row-of-dataframe-with-multiple-argum
sr_unadj_noE_col <- apply(solowResiduals[ , c("Country", "model", "energy", "Year", "flavor")], 1, function(x){
    row <- as.data.frame(filter(solowResiduals, 
                                Country==x["Country"], model==x["model"], energy=="Without energy",
                                Year==x["Year"], flavor=="Unadjusted"))
    return(row[["lambda_fd"]])
  })
solowResiduals[["sr_unadj_noE"]] <- do.call(rbind, as.list(sr_unadj_noE_col))
# Create a column containing ∆lambda_df, the difference between a lambda_df value
# and the lambda_df value in the same year for the unadjusted, no-energy case
solowResiduals[["Dlambda_fd"]] <- abs(solowResiduals[["lambda_fd"]]) - abs(solowResiduals[["sr_unadj_noE"]])

# Adjust levels before creating graphs.
solowResiduals$flavor <- relevelFactor(as.factor(solowResiduals$flavor), c("Unadjusted", "Quality-adjusted"))
solowResiduals$energy <- relevelFactor(as.factor(solowResiduals$energy), c("Without energy", "With energy"))
@

@@

<<DynamicSR_CD_Graph, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Dynamic Solow residuals for the Cobb-Douglas model. Dashed lines are $\\lambda$ values. Solid lines are $\\frac{1}{y}\\frac{\\mathrm{d} y}{\\mathrm{d} t}$.">>=
data <- filter(solowResiduals, model=="CD")
ggplot(data=data) +
  geom_line(data=filter(data, energy=="Without energy"), mapping=aes(x=Year, y=dydtOvery)) +
  geom_line(mapping=aes(x=Year, y=lambda_fd, linetype=energy)) +
  scale_linetype_manual(values=c(2,3), name="") +
  facet_grid(Country ~ flavor) +
  scale_x_continuous(breaks=c(1960, 1980, 2000)) +
  xlab("") + 
  ylab(expression(lambda[ij])) +
  xy_theme()
@

%++++++++++++++++++++++++++++++
\subsection{How does quality-adjusting and including energy affect the dynamic Solow residual?} 
\label{sec:trajectory_graphs}
%++++++++++++++++++++++++++++++

<<SR and R trajectory graph functions>>=
residualsGraph <- function(data){
  plot <- ggplot(data=data) + 
    geom_point(mapping=aes(x=Dr, y=Dlambda_fd)) + 
    xlab(expression(Delta*r)) +
    ylab(expression(Delta*lambda)) +
    xy_theme()
  return(plot)
}
traj.data <- merge(solowResiduals[c("Country", "Year", "model", "flavor", "energy", "nest", "Dlambda_fd")], 
                   fitted_and_resid_data[c("Country", "Year", "model", "flavor", "energy", "nest", "Dr")])
traj.data.means <- traj.data %>% 
  group_by(model, Country, flavor, energy, nest) %>%
  summarise(Dlambda_fd.mean = mean(Dlambda_fd, na.rm=TRUE), Dr.mean = mean(Dr))
@

We can test our hypothesis that quality-adjusting the factors of production
and including energy will reduce the Solow residual ($\lambda$)
and the fitting residuals ($r$) by plotting
$\Delta\lambda$ vs. $\Delta r$, where
%
\begin{equation} \label{eq:delta_lambda}
  \Delta \lambda \equiv \lambda - \lambda_{Unadjusted, \, Without \, energy}
\end{equation}
%
and
%
\begin{equation} \label{eq:delta_r}
  \Delta r \equiv r - r_{Unadjusted, \, Without \, energy}.
\end{equation}
%

<<Solow_residual_trajectories, fig.pos="H", fig.align='center', fig.width=6.5, fig.height=4, fig.cap="Dynamic Solow residual trajectories for the Cobb-Douglas model.">>=
residualsGraph(subset(traj.data, model=="CD" & Country=="UK" & energy=="With energy" & flavor=="Unadjusted"))
@
